---
title: "Overeducation Subgroup Analysis and EDA"
author: "Alisha Gurnani"
date: "5/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set()


#opts_knit$set(root.dir = "Thesis/Data/Processed")
library(tidyverse)
library(dplyr)
library(tidycensus)
library(gdata)
library(ggplot2)
library(stargazer)
library(gsubfn)
library(plotly)
library(tidyr)
library(lemon)
library(viridis)
```

## Overeducation Exploratory Analysis

###  Read in all datasets

```{r}
#read in all of the separate CSVS
setwd("Thesis/Data/Processed")

#read in all IPUMS ACS data first
ACS_2004 <- read_csv("ACS_2004.csv")%>%
 select(YEAR, OCC, OCCSOC, STATEFIP, AGE, RACE, SEX, MARST, CITIZEN, INCTOT, REGION,IND, EducAttainment, Overeducation)
ACS_2005 <- read_csv("ACS_2005.csv")%>%
 select(YEAR, OCC, OCCSOC, STATEFIP, AGE, RACE, SEX, MARST, CITIZEN, INCTOT, REGION,IND, EducAttainment, Overeducation)

ACS_2006 <-read_csv("ACS_2006.csv")%>%
  select(YEAR, OCC, OCCSOC, STATEFIP, AGE, RACE, SEX, MARST, CITIZEN, INCTOT, REGION,IND, EducAttainment, Overeducation)

ACS_2007 <-read_csv("ACS_2007.csv")%>%
  select(YEAR, OCC, OCCSOC, STATEFIP, AGE, RACE, SEX, MARST, CITIZEN, INCTOT, REGION,IND, EducAttainment, Overeducation)
ACS_2008 <- read_csv("ACS_2008.csv")%>%
  select(YEAR, OCC, OCCSOC, STATEFIP, AGE, RACE, SEX, MARST, CITIZEN, INCTOT, REGION,IND, EducAttainment, Overeducation)
ACS_2009 <- read_csv("ACS_2009.csv")%>%
  select(YEAR, OCC, OCCSOC, STATEFIP, AGE, RACE, SEX, MARST, CITIZEN, INCTOT, REGION,IND, EducAttainment, Overeducation)
ACS_2010 <- read_csv("ACS_2010.csv")%>%
  select(YEAR, OCC, OCCSOC, STATEFIP, AGE, RACE, SEX, MARST, CITIZEN, INCTOT, REGION,IND, EducAttainment, Overeducation)
ACS_2011 <- read_csv("ACS_2011.csv")%>%
 select(YEAR, OCC, OCCSOC, STATEFIP, AGE, RACE, SEX, MARST, CITIZEN, INCTOT, REGION,IND, EducAttainment, Overeducation)
ACS_2012 <- read_csv("ACS_2012.csv")%>%
 select(YEAR, OCC, OCCSOC, STATEFIP, AGE, RACE, SEX, MARST, CITIZEN, INCTOT, REGION,IND, EducAttainment, Overeducation)

#read in BLS GDP 2004-2012
BLS_GDP <- read_csv("Thesis/Data/Processed/BeaGDP.csv") %>%
  select(TimePeriod, DataValue)  %>%
  rename(YEAR = TimePeriod,
         GDP = DataValue) %>%
  filter(YEAR != 2013)
```


```{r}
#view BLS
BLS_GDP
```

```{r}
#plot GDP by year - line graph
GDP_Plot<- ggplot(data=BLS_GDP, aes(x=YEAR, y=GDP, group=1)) +
  geom_line()+
  geom_point() +
  labs(title="U.S. Gross Domestic Product by Year",x="Year", y = "GDP")

```


```{r}
### join all datasets

All_ACS <- do.call("rbind", list(ACS_2004, ACS_2005, ACS_2006, ACS_2007, ACS_2008, ACS_2009, ACS_2010, ACS_2011, ACS_2012))

```

```{r}
#view all acs dataset
head(All_ACS)
```



## breakdown of overeducation and non-overeducated per year 
```{r}
#Overeducation data with NAS
Overeducation_NAS <- All_ACS %>%
  group_by(YEAR,Overeducation) %>%
  dplyr::summarise(Number=n()) %>%
   mutate(percent = Number / sum(Number) * 100)

head(Overeducation_NAS)
```


```{r}
#Overeducation data without NAS

Overeducation_NO_NAS <- All_ACS  %>%
  filter(!is.na(Overeducation))%>%
  group_by(YEAR,Overeducation) %>%
  dplyr::summarise(Number=n()) %>%
  mutate(percent = Number / sum(Number) * 100)
head(Overeducation_NO_NAS)

```

## distribution of Race through the data
```{r}
#distribution of Race
Race1 <- All_ACS %>%
  filter(!is.na(Overeducation))%>%
  group_by(YEAR, RACE)%>%
  summarise(Count = n()) %>%
  mutate(percent = Count / sum(Count) * 100)
head(Race1)
```



## Find Percentages per each variable : Region and Age



### distribution of age over time
```{r}

age_data<- All_ACS %>%
  filter(!is.na(Overeducation))%>%
  select(YEAR, AGE, Overeducation)
```

```{r}
str(All_ACS$AGE)
```



```{r}
age_data["Age_Groups"] <- ifelse(age_data$AGE>=16 & age_data$AGE<=24, "16 to 24 years",
                                      ifelse(age_data$AGE >= 25 & age_data$AGE <= 34, "25 to 35 years",
                                             ifelse(age_data$AGE >= 35 & age_data$AGE <= 44, "35 to 44 years",
                                                    ifelse(age_data$AGE >= 45 & age_data$AGE <= 54, "45 to 54 years",
                                                           ifelse(age_data$AGE >= 55 & age_data$AGE <= 64 , "55 to 64 years",
                                                                  ifelse(age_data$AGE >= 65 , "65 years and Up", NA))))))
```


```{r}
#AGE -- create buckets for age first - or line graph with x=age and y=percent
Age <- age_data %>%
  filter(!is.na(Overeducation))%>%
  filter(!is.na(Age_Groups)) %>%
  group_by(YEAR, Age_Groups, Overeducation)%>%
  summarise(Count = n())

head(Age)

```


```{r}
#age and rate of overeducation per year
Overeducation_Age <- age_data %>%
  filter(!is.na(Overeducation))%>%
  filter(!is.na(Age_Groups)) %>%
  group_by(YEAR, Age_Groups, Overeducation)%>%
  dplyr::summarise(Count=n()) %>%
  mutate(percent = Count / sum(Count) * 100)

head(Overeducation_Age)
```


```{r}
#plot for age and rate of overeducation over time
age_barplot <- Overeducation_Age %>%
  filter(Overeducation != 0) %>%
  filter(!is.na(Age_Groups)) %>%
  ggplot(
    aes(x=Age_Groups, y = percent/100, fill = Age_Groups)) +
  geom_bar(stat = "identity")+
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Percent of Overeducation per Age Group by Year", x= "Age", y = "Percent of Overeducation", fill = "Age Groups") +
  theme(axis.text.x = element_blank(), axis.ticks = element_blank())+ 
  facet_wrap(~YEAR)
```



### distribution of population in different regions overtime

```{r}
Regions_data <- All_ACS %>%
  filter(!is.na(Overeducation))%>%
  select(YEAR, REGION, Overeducation, STATEFIP)
```

```{r}
str(Regions_data$STATEFIP)
```


The regions will be set based on the BEA regional definition as follows: Far West, Great Lakes, Mideast, New England, Plains, Rocky Mountain, Southeast, and Southwest. 
Far West:
  - California - 06
  - Nevada - 32
  - Oregon - 41
  - Washington - 53
  -Alaska - 2
  -hawaii -15
Great Lakes:
  - Wisconsan - 55
  - Illinois - 17
  - Michigan - 26
  - Indiana - 18
  - Ohio - 39
Mideast:
  -New York - 36
  -Pennsylvania - 42
  - New Jersey - 34
  - Deleware - 10
  - Maryland - 24
  - D.C --??
New England:
  -Maine - 23
  -Massachusetts - 25
  -New Hampshire - 33
  -Rhode Island - 44
  - Connecticut - 9
  -Vermont - 50
Plains:
  -North Dakota - 38
  -South Dakota - 46
  - Minnesota - 27
  - Nebraska - 31
  - Kansas - 20 
  - Missouri -29
  - Iowa -19
Rocky Mountains:
  -Montana - 30
  - Idaho -16
  - Wyoming - 56
  -Utah - 49
  - Colorado -8
 Southeast:
  -Virginia - 51
  -West Virginia - 54
  - North Carolina - 37
  - South Carolina - 45
  - Georgia - 13
  -Florida - 12
  -Kentucky - 21
  -Tennessee - 47
  -Alabama - 1
  -Mississippi - 28
  -Arkansas - 5
  -Louisiana- 22
 Southwest:
  - arizona - 4
  - New Mexico - 35
  -Oklahoma - 40
  - Texas - 48


```{r}
Regions_data["Census_Region"] <- NA
Regions_data$Census_Region[Regions_data$STATEFIP ==  6 | Regions_data$STATEFIP == 32 | Regions_data$STATEFIP== 41| Regions_data$STATEFIP== 53| Regions_data$STATEFIP== 2|
                             Regions_data$STATEFIP==15] = "Far West" 
Regions_data$Census_Region[Regions_data$STATEFIP == 17| Regions_data$STATEFIP == 18| Regions_data$STATEFIP ==26| Regions_data$STATEFIP == 39| Regions_data$STATEFIP == 55] = "Great Lakes"
Regions_data$Census_Region[Regions_data$STATEFIP == 10| Regions_data$STATEFIP == 24| Regions_data$STATEFIP == 34| Regions_data$STATEFIP ==36| Regions_data$STATEFIP ==42] = "Mideast"
Regions_data$Census_Region[Regions_data$STATEFIP == 9| Regions_data$STATEFIP == 23| Regions_data$STATEFIP == 25| Regions_data$STATEFIP == 33| Regions_data$STATEFIP == 44|
                             Regions_data$STATEFIP == 50] = "New England"
Regions_data$Census_Region[Regions_data$STATEFIP == 19| Regions_data$STATEFIP == 20| Regions_data$STATEFIP == 29| Regions_data$STATEFIP == 27| Regions_data$STATEFIP == 31|
                             Regions_data$STATEFIP == 38| Regions_data$STATEFIP == 46] = "Plains"
Regions_data$Census_Region[Regions_data$STATEFIP== 8| Regions_data$STATEFIP == 16| Regions_data$STATEFIP == 30| Regions_data$STATEFIP == 49| Regions_data$STATEFIP == 56] = "Rocky Mountains"
Regions_data$Census_Region[Regions_data$STATEFIP== 1| Regions_data$STATEFIP == 5| Regions_data$STATEFIP == 12| Regions_data$STATEFIP == 21| Regions_data$STATEFIP == 13| 
                             Regions_data$STATEFIP== 22| Regions_data$STATEFIP == 28| Regions_data$STATEFIP ==37| Regions_data$STATEFIP == 45| Regions_data$STATEFIP == 47| 
                             Regions_data$STATEFIP == 51| Regions_data$STATEFIP == 54] = "Southeast"
Regions_data$Census_Region[Regions_data$STATEFIP== 4| Regions_data$STATEFIP ==35| Regions_data$STATEFIP == 40| Regions_data$STATEFIP == 48] = "Southwest"
```


```{r}
#Region data
Regions <- Regions_data %>%
  filter(!is.na(Overeducation))%>%
  filter(!is.na(Census_Region)) %>%
  group_by(YEAR, Census_Region, Overeducation)%>%
  summarise(Count = n())

head(Regions)

```


```{r}
#region and rate of overeducation per year
Overeducation_Region <- Regions_data %>%
  filter(!is.na(Overeducation))%>%
  filter(!is.na(Census_Region)) %>%
  group_by(YEAR, Census_Region, Overeducation)%>%
  dplyr::summarise(Count=n()) %>%
  mutate(percent = Count / sum(Count) * 100)

head(Overeducation_Region)
```


```{r}
#plot for region and rate of overeducation over time
regions_barplot <- Overeducation_Region %>%
  filter(Overeducation != 0) %>%
  filter(!is.na(Census_Region)) %>%
  ggplot(
    aes(x=Census_Region, y = percent/100, fill = Census_Region)) +
  geom_bar(stat = "identity")+
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Percent of Overeducation per Region in the United States by Year", x= "Region", y = "Percent of Overeducation", fill = "Region") +
  theme(axis.text.x = element_blank(), axis.ticks = element_blank())+ 
  facet_wrap(~YEAR)+
  scale_color_viridis(discrete = TRUE, option = "D")+
  scale_fill_viridis(discrete = TRUE) 
  
```


